Allocation scores, WIS, and CRPS via decision theory

### 2023-09-11 alloscore_manuscript/methods_apg.tex
AS approx idea start
Talk about KKT.

The optimal allocation $x^{F,k}$ according to a forecast $F$ is a solution to the constrained optimization, or \emph{allocation} problem
\begin{align}
    (\mathrm{AP}) \quad \underset{x \in \mathbb{R}_{+}^N}{\mathrm{minimize}}\,\, E_{F}\left[s_A(Y,x)\right] \text{ subject to } 
    T_x = k. \label{AP}
\end{align}
The AP has a convex objective function and affine constraint, so from general convex optimization theory 
(see e.g. \cite{ruszczynski2011nonlinear}, Chapter 4), it is equivalent to the
problem of finding a saddle point $(x^{F,k}, \lambda^{\star})$ of the associated Lagrangian 
\begin{align}
\mathcal{L}(x, \lambda) = \mathcal{L}(x, \lambda; F, k, L) 
&= E_F\left[s_A(Y,x)\right] + \lambda(T_x - k) \\
&= L\sum_{i=1}^{N} E_{F_i}[(Y_i - x_i)_{+}] + \lambda\left(\sum_{i=1}^{N} x_i - k\right).  
\end{align}
in the region $\{x \geq 0, \lambda \in \mathbb{R}\}$.\footnote{The Lagrangian in this context has the interpretation as the best (lowest) objective function value we can achieve if we forced to choose $x$ but can buy or get credit for changes the constraint level $k$ 
at the ``price'' $\lambda$.}
The Lagrangian reformulation in particular recasts the constraint $T_x=k$ of the original problem as a first-order condition on $\mathcal{L}$ with respect to the new \emph{Lagrange multiplier} variable $\lambda$:
\begin{align}
0 = \del{}{\lambda} \mathcal{L}(x^{F,k}, \lambda^{\star}) &= \sum_{i=1}^{N} x_i^{F,k} - k. \label{eqn:lambda_stationary}
\end{align}
But for $(x^{F,k}, \lambda^{\star})$ to be a saddle point we also need
\begin{align}
0 = \frac{\partial}{\partial x_i} \mathcal{L}(x^{F.k}, \lambda) &= L(F_i(x_i^{F,k})-1) + \lambda^{\star} 
\quad (\text{cf. \eqref{eqn:q_deriv}}) \label{eqn:x_stationary}
\end{align}
for any $x_i^{F,k} > 0$, which gives a formula 
\begin{align}
x_i^{F,k} = F_i^{-1}(1-\lambda^{\star}/L) \label{eqn:xfk_formula}
\end{align}
for any non-zero element $x_i^{F,k}$ of the solution. Thus \eqref{eqn:lambda_stationary} becomes
\begin{align}
0 = \sum_{\{i | x_i^{F,k} > 0\}} F_{i}^{-1}(1-\lambda^{\star}/L) - k. \label{eqn:lambda_stationary2}
\end{align}

Since convexity and compactness guarantee existence of a solution, the problem is now reduced to solving \eqref{eqn:lambda_stationary} for the set 
$\{i \mid x_i^{F,k} > 0\}$ and value $\lambda^{\star}$. Note, however, that this is a more complicated task than solving just 
for $\lambda^{\star}$ under the assumption that \eqref{eqn:xfk_formula} holds for all coordinates of $x^{F,k}$.\footnote{Such an assumption would exclude fairly simple examples in which $F$ has support bounded away from $0$ in one coordinate but not in another.} The task is simplified though by bringing in the full saddle point condition for $\lambda^{\star}$ which implies not only \eqref{eqn:lambda_stationary} but also that
$\lambda^{\star}$ is in fact the unique maximizer of the \emph{dual function} 
$\mathcal{L}_D(\lambda) := \inf_{x \in \mathbb{R}_{+}^N} \mathcal{L}(x, \lambda)$.
This approach is especially suitable in our situation because our original objective function $E_F\left[s_A(Y,x)\right]$ is decomposable in the coodinates $x_i$, allowing the dual function to be computed separately with respect to each coordinate:
\begin{align}
\mathcal{L}_D(\lambda) &= \inf_{x \in \mathbb{R}_{+}^N} L\sum_{i=1}^{N} E_{F_i}[(Y_i - x_i)_{+}] + \lambda\left(\sum_{i=1}^{N} x_i - k\right) \\
&= - \lambda k + \sum_{i=1}^{N} \inf_{x_i \in \mathbb{R}_{+}}\left\{ L E_{F_i}[(Y_i - x_i)_{+}] + \lambda x_i\right\}  \label{eqn:dual_decomp} \\
&= - \lambda k + \sum_{i=1}^{N}  LE_{F_i}[(Y_i - x_i(\lambda))_{+}] + \lambda x_i(\lambda).
\end{align}
Here, $x_i(\lambda)$ is any member of the set 
\begin{align}
\chi_i(\lambda) = \underset{x_i \in \mathbb{R}_{+}}{\mathrm{arginf}} \left\{ L E_{F_i}[(Y_i - x_i)_{+}] + \lambda x_i\right\}
\end{align}
which for $F_i$ in a standard parametric family will often have simple the form $\chi_i(\lambda) = \{F_i^{-1}(1-\lambda/L)\}$ but could an interval containing $0$ or $\infty$.
From general theory, $\mathcal{L}_D(\lambda)$ is concave, and while it may not be differentiable when the $x_i$'s attaining the infima in \eqref{eqn:dual_decomp} are zero, it does have as a subdifferential the well-defined interval
\begin{align}
\partial \mathcal{L}_D(\lambda) 
&= \sum_{i=1}^N \chi_i(\lambda) - k.
\end{align}
This of course looks very similar to the right hand side of \eqref{eqn:lambda_stationary2}, but the dual perspective has the key advantage of not requiring any guesses of the set $\{i \mid x_i^{F,k} > 0\}$ when initiating a search algorithm for $\lambda^{\star}$.  Instead, given a candidate $\lambda_0$ for $\lambda^{\star}$, we proceed by calculating the sets $\chi_i(\lambda_0), i = 1,\ldots,N$, and using them to find the (usually zero-width) interval $\partial \mathcal{L}_D(\lambda_0)$. If $\partial \mathcal{L}_D(\lambda_0) \subset (-\infty,0)$, we can eliminate $\{\lambda \geq \lambda_0\}$ from our search and similarly if $\partial \mathcal{L}_D(\lambda_0) \subset (0, \infty)$, we can eliminate $\{\lambda \leq \lambda_0\}$.

@@

>>>>

The optimal allocation $x^{F,k}$ according to a forecast $F$ is a solution to the constrained optimization, or \emph{allocation} problem
\begin{align}
    (\mathrm{AP}) \quad \underset{x \in \mathbb{R}_{+}^N}{\mathrm{minimize}}\,\, E_{F}\left[s_A(Y,x)\right] \text{ subject to } 
    T_x = k. \label{AP}
\end{align}
The AP has a convex objective function and affine constraint, so from general convex optimization theory, a solution $x^{F,k}$ must satisfy a set of \emph{Karush-Kuhn-Tucker} (KKT) conditions. These include that for any $i$ such that $x_i^{F,K} > 0$,
\begin{align}
\frac{\partial}{\partial x_i} E_{F}\left[s_A(Y,x^{F,k})\right] &= L(F_i(x_i^{F,k})-1) = \lambda
\quad (\text{cf. \eqref{eqn:q_deriv}}) \label{eqn:x_stationary}
\end{align}
for some $\lambda$ independent of $i$. These equations follow from the assumption that when $x=x^{F,k}$, the objective $E_{F}\left[s_A(Y,x)\right]$ cannot be reduced by any small
resource shifts between locations and they provide a quantile formula for any non-zero optimal allocation level,
\begin{align}
x_i^{F,k} = F_i^{-1}(1-\lambda/L). \label{eqn:xfk_formula}
\end{align}
Another KKT condition is the constraint equation, which can now be written in terms of $\lambda$ as
\begin{align}
0 = \sum_{i \in I_+} F_{i}^{-1}(1-\lambda/L) - k, \label{eqn:lambda_stationary2}
\end{align}
where $I_+ = \{i \mid x_i^{F,k} > 0\}$.
Since convexity and compactness guarantee existence of a solution, the problem is now reduced to solving 
\eqref{eqn:lambda_stationary2} for the set 
$\{i \mid x_i^{F,k} > 0\}$ and a root $\lambda^{\star}$, in terms of which we can exrpess the Bayes act as
\begin{align}
x_i^{F,k} = 
\begin{cases}
F_i^{-1}(1-\lambda^{\star}/L) &  i \in I_+ \\
0 & \text{otherwise}.
\end{cases}
\end{align}
Note, however, that this is a more complicated task than just solving \eqref{eqn:lambda_stationary2}
for $\lambda^{\star}$ under the assumption that $I_+ = \{1,\ldots,N\}$.\footnote{Such an assumption would exclude fairly simple examples in which $F$ has support bounded away from $0$ in one coordinate but not in another.}

The task can be simplified though by turning our attention to the \emph{dual} formulation of the AP. (See e.g. \cite{ruszczynski2011nonlinear}, Chapter 4, which we follow here.) This involves recasting the KKT equations \eqref{eqn:x_stationary}
and the constraint $T_x=k$ as the first order conditions in variables $x$ and $\lambda$, respectively, for $(x^{F,k}, \lambda^{\star})$ to be a saddle point of the \emph{Lagrangian}
\begin{align}{}
\mathcal{L}(x, \lambda) = \mathcal{L}(x, \lambda; F, k, L) 
&= E_F\left[s_A(Y,x)\right] + \lambda(T_x - k) \\
&= L\sum_{i=1}^{N} E_{F_i}[(Y_i - x_i)_{+}] + \lambda\left(\sum_{i=1}^{N} x_i - k\right).  
\end{align}
in the region $\{x \geq 0, \lambda \in \mathbb{R}\}$.\footnote{The Lagrangian in this context has the interpretation as the best (lowest) objective function value we can achieve if we forced to choose $x$ but can buy or get credit for changes in the constraint level $k$ 
at the ``price'' $\lambda$.}
The \emph{dual function} is then defined as $\mathcal{L}_D(\lambda) := \inf_{x \in \mathbb{R}_{+}^N} \mathcal{L}(x, \lambda)$. 
From general theory, $\mathcal{L}_D(\lambda)$ is concave and $\lambda^{\star}$ is its unique maximizer.
The reason why the dual approach is especially suitable in our situation is that 
our original objective function $E_F\left[s_A(Y,x)\right]$ is decomposable in the coodinates $x_i$, allowing the corresponding infima in $\mathcal{L}_D$ to be computed separately:
\begin{align}
\mathcal{L}_D(\lambda) &= \inf_{x \in \mathbb{R}_{+}^N} L\sum_{i=1}^{N} E_{F_i}[(Y_i - x_i)_{+}] + \lambda\left(\sum_{i=1}^{N} x_i - k\right) \\
&= - \lambda k + \sum_{i=1}^{N} \inf_{x_i \in \mathbb{R}_{+}}\left\{ L E_{F_i}[(Y_i - x_i)_{+}] + \lambda x_i\right\}  \label{eqn:dual_decomp} \\
&= - \lambda k + \sum_{i=1}^{N}  LE_{F_i}[(Y_i - x_i(\lambda))_{+}] + \lambda x_i(\lambda).
\end{align}
Here, $x_i(\lambda)$ is any member of the set 
\begin{align}
\chi_i(\lambda) = \underset{x_i \in \mathbb{R}_{+}}{\mathrm{arginf}} \left\{ L E_{F_i}[(Y_i - x_i)_{+}] + \lambda x_i\right\}
\end{align}
which for $F_i$ in a standard parametric family will often simply be $\{F_i^{-1}(1-\lambda/L)\}$ but could an interval containing $0$ or $\infty$.
While $\mathcal{L}_D(\lambda)$ may not be differentiable when the $x_i$'s attaining the infima in \eqref{eqn:dual_decomp} are zero, it does have as a well-defined subdifferential which is the sum of intervals
\begin{align}
\partial \mathcal{L}_D(\lambda) 
&= \sum_{i=1}^N \chi_i(\lambda) - k \\
&= \left[ \sum_{i=1}^N \min(\chi_i(\lambda))  - k, \sum_{i=1}^N \max(\chi_i(\lambda))  - k\right]
\end{align}

% The Lagrangian reformulation in particular recasts the constraint $T_x=k$ of the original problem as a first-order condition on $\mathcal{L}$ with respect to the new \emph{Lagrange multiplier} variable $\lambda$:
% \begin{align}
% 0 = \del{}{\lambda} \mathcal{L}(x^{F,k}, \lambda^{\star}) &= \sum_{i=1}^{N} x_i^{F,k} - k. \label{eqn:lambda_stationary}
% \end{align}
% But for $(x^{F,k}, \lambda^{\star})$ to be a saddle point we also need
% \begin{align}
% 0 = \frac{\partial}{\partial x_i} \mathcal{L}(x^{F.k}, \lambda) &= L(F_i(x_i^{F,k})-1) + \lambda^{\star} 
% \quad (\text{cf. \eqref{eqn:q_deriv}}) \label{eqn:x_stationary}
% \end{align}
% for any $x_i^{F,k} > 0$, which gives a formula 
% \begin{align}
% x_i^{F,k} = F_i^{-1}(1-\lambda^{\star}/L) \label{eqn:xfk_formula}
% \end{align}
% for any non-zero element $x_i^{F,k}$ of the solution. Thus \eqref{eqn:lambda_stationary} becomes
% \begin{align}
% 0 = \sum_{\{i | x_i^{F,k} > 0\}} F_{i}^{-1}(1-\lambda^{\star}/L) - k. \label{eqn:lambda_stationary2}
% \end{align}

This of course looks very similar to the right hand side of \eqref{eqn:lambda_stationary2}, but the dual perspective has the key advantage of not requiring any guesses of the set $\{i \mid x_i^{F,k} > 0\}$ when initiating a search algorithm for $\lambda^{\star}$.  Instead, given a candidate $\lambda_0$ for $\lambda^{\star}$, we proceed by calculating the sets $\chi_i(\lambda_0), i = 1,\ldots,N$, and using them to find the (usually zero-width) interval $\partial \mathcal{L}_D(\lambda_0)$. If $\partial \mathcal{L}_D(\lambda_0) \subset (-\infty,0)$, we can eliminate $\{\lambda \geq \lambda_0\}$ from our search and similarly if $\partial \mathcal{L}_D(\lambda_0) \subset (0, \infty)$, we can eliminate $\{\lambda \leq \lambda_0\}$.

@@

\section{Approximating the allocation score}

cdf of $TY_F$:
\begin{align}
TF(K) = \sup \left\{\tau \in [0,1] \mid \sum F_i^{-1}(\tau) \leq K \right\} 
\end{align}
Given a set of distinct and increasing probability levels $\boldsymbol{\alpha} = \{\alpha_1,\ldots,\alpha_J\} \subset (0,1)$, we can define a function
\begin{align}
j_{\boldsymbol{\alpha},-}(K) = \max \left\{j \in \{1,\ldots,J\} \mid \sum F_i^{-1}(\alpha_j) \leq K \right\} 
\end{align}

An approximation of $S_{A,K}(F,y)$ using elicited quantiles:
\begin{align}
S_{A, \boldsymbol{\alpha}}(F,y) := w_1 s_A(q_{F,j}, y) + w_2 s_A(q_{F,j+1}, y)
\end{align}
